Effectful Semantics for Substructural SSA

### 2023-11-16 pldi24/paper.tex
Tap at rewriting appendix section

\begin{figure}
  \begin{center}        
    \begingroup
    \renewcommand{\arraystretch}{1.5}
    \setlength{\tabcolsep}{2em}
    \begin{tabular}{rl}
        \multicolumn{1}{c}{Judgment} & \multicolumn{1}{c}{Meaning} \\ \hline
        \(\issubst{\gamma}{\Theta}{\Gamma}\) &
        \(\gamma\) is a substitution taking \(\Theta\) to \(\Gamma\) \\
        \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\) &
        \(\mc{L}\) is a label substitution taking \(\ms{L}\) to \(\ms{K}\) \\
        % \(\mhasty{H}{\Gamma}{p}{\mhole{a}}{A}{q}\) &
        % ... \\
        % \(\mhaslb{H}{\Gamma}{\mhole{t}}{\ms{L}}\) &
        % ... \\
        % \(\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}}\) &
        % ... \\
        % \(\isrw{\mc{H}}{I}{H}\) &
        % ... \\
    \end{tabular}
    \endgroup
  \end{center}
  \caption{Metatheoretic typing judgements for \isotopessa}
  \Description{Metatheoretic typing judgements for isotope-SSA}
  \label{fig:ssa-meta-judgements}
\end{figure}

@@

We begin with some basic definitions:
\begin{definition}[Binoidal Category] 
  A \textbf{binoidal category} \(\mc{C}\) is a category equipped with a binary
  operation on objects \(\otimes: \mc{C} \times \mc{C} \to \mc{C}\) which is
  independently functorial in each argument (i.e., for each object \(A\), \(A
  \otimes -\) and \(- \otimes A\) are functors).
\end{definition}
Given \(f: \mc{C}(A, B)\), \(g: \mc{C}(A', B')\) in a binoidal category, we
write \(f \ltimes g = f \otimes A';B \otimes g\) and \(f \rtimes g = A \otimes
g;f \otimes B'\); we say \(f\) is \textbf{central} if, for all \(g\), \(f
\ltimes g = f \rtimes g\) and \(g \ltimes f = g \rtimes f\), in which case we
simply write \(f \otimes g\), \(g \otimes f\) respectively.
\begin{definition}[Premonoidal Category] 
  A \textbf{premonoidal category} is a binoidal category with an identity object
  \(I: \mc{C}\) such that
  \begin{itemize}
    \item There exist central natural isomorphisms \(\lambda_A: I \otimes A
    \simeq A\) (the \textbf{left unitor}), \(\rho_A: A \otimes I \simeq A\) (the
    \textbf{right unitor})
    \item There exists a central isomorphism \(\alpha_{A, B, C}: (A \otimes B)
    \otimes C \simeq A \otimes (B \otimes C)\) (the \textbf{associator})
    \item Satisfying the \textbf{pentagon} and \textbf{triangle laws}; see e.g.
    \citet{maclane:71}.
  \end{itemize}
  We say a premonoidal category is \textbf{symmetric} if it is equipped with a
  central natural isomorphism \(\sigma_{A, B}: A \otimes B \simeq B \otimes A\)
  (the \textbf{symmetry}) satisfying the \textbf{hexagon law} (see
  \cite{maclane:71}). We say a premonoidal category is \textbf{monoidal} if
  every morphism is central.
\end{definition}
One issue with the notion of a premonoidal category is that it is unclear what
exactly a premonoidal functor should be: if we want them to compose, we would
like to preserve central morphisms, but trying to preserve \textit{all} central
morphisms is too strict a notion. We want, in essence, to be able to distinguish
morphisms that are truly "pure" from those which simply happen to be central. To
do that, we introduce the notion of an \textit{effectful category} as follows:
\begin{definition}[Effectful Category] 
  An \textbf{effectful category} \(\mc{V} \to \mc{C}\) is an identity on objects
  functor \(\upg{\cdot}{}: \mc{V} \to \mc{C}\) from a monoidal category
  \(\mc{V}\) to a premonoidal category \(\mc{C}\) which preserves all
  premonoidal structure (i.e. sends associators to associators, unitors to
  unitors, and tensor products to tensor products). It is \textbf{symmetric} if
  both \(\mc{V}\) and \(\mc{C}\) are, and \(\upg{\cdot}{}\) preserves the
  symmetry. A \textbf{Freyd category} is an effectful category where \(\mc{V}\)
  is Cartesian.
\end{definition}
Unfortunately, the notion of "purity" given by an effectful category only allows
us to perform relatively weak optimizations in the presence of effects. To
remedy this, we want to be able to make a finer distinction between morphisms
which are merely \textit{central} and those which are also \textit{relevant}
(i.e., copying the input and applying them twice is the same as applying them
once and copying the output) and/or \textit{affine} (i.e., applying them and
discarding the result is the same as never doing them at all). Doing so also
allows us to add the capability of \textit{distinguishing} between objects which
are affine, relevant, neither or both, enabling us to reason about substructural type systems categorically.

@@

Substitution only applies to central terms, because those are the only terms for
which substitution is sound. Furthermore, there are many other optimisations
that follow from the beta-eta theory of the language, as well as specific
properties of model-dependent operations (such as memory operations). In the
appendix, we prove a general rewriting theorem which lifts equations between
terms into a full congruence. This lets us justify optimisations such as scalar
replacement of aggregates, eliminating redundant writes, reordering independent
memory operations, and memory-to-register transformations.

@@

\section{Proofs}

\begin{theorem}[Weakening] \
  \begin{itemize}
    \item If \(\cwk{\Gamma}{\Delta}\) and \(q \subseteq r\),
    \(\hasty{\Delta}{p}{a}{A}{r}\), then \(\hasty{\Gamma}{p}{a}{A}{q}\)
    \item If \(\cwk{\Gamma}{\Delta}\), \(\haslb{\Delta}{t}{\ms{L}}\), then
    \(\haslb{\Gamma}{t}{\ms{L}}\)
    \item If \(\lwk{\ms{L}}{\ms{K}}\), \(\haslb{\Gamma}{t}{\ms{L}}\) then
    \(\haslb{\Gamma}{t}{\ms{K}}\)
  \end{itemize}
\end{theorem}

\begin{proof} \
  \TODO{this}
\end{proof}

\begin{lemma}[Substitution Splitting] 
  If \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\issubst{\gamma}{\Theta}{\Gamma}\),
  then there exist \(\Theta_\Delta\), \(\Theta_\Xi\) such that:
  \begin{itemize}
    \item \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\)
    \item \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\)
    \item \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\)
  \end{itemize}
\end{lemma}

\begin{proof} 
  We proceed by induction on derivations \(\csplits{\Gamma}{\Delta}{\Xi}\). In
  the \rle{split-nil} case, we have that \(\issubst{\gamma}{\cdot}{\cdot}\) by
  \rle{subst-nil}. For all other cases, since \(\issubst{\gamma}{\Theta}{\Gamma,
  \thyp{x}{A}{q}}\), we have by \rle{subst-cons} that
  \(\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}\),
  \(\issubst{\gamma}{\Theta_\Gamma}{\Gamma}\),
  \(\hasty{\Theta_\Gamma}{1}{\gamma(x)}{A}{q}\). Since
  \(\csplits{\Gamma}{\Delta}{\Xi}\), by induction, we have that
  \(\csplits{\Theta_\Gamma}{\Theta_\Delta}{\Theta_\Xi}\) where \(\Theta_\Delta\)
  and \(\Theta_\Xi\) have the desired properties above.
  \begin{itemize}
    \item \rle{split-left}: by left split association, we have that there exists
    \(\Theta_{\Delta, \thyp{x}{A}{q}}\), \(\csplits{\Theta}{\Theta_{\Delta,
    \thyp{x}{A}{q}}}{\Theta_\Xi}\), \(\csplits{\Theta_{\Delta,
    \thyp{x}{A}{q}}}{\Theta_\Delta}{\Theta_x}\). Using the fact that
    \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\), we can conclude by
    \rle{subst-cons} that \(\issubst{\gamma}{\Theta_{\Delta,
    \thyp{x}{A}{q}}}{\Delta, \thyp{x}{A}{q}}\); the desired result hence
    follows.
    \item \rle{split-right}: by right split association, we have that there
    exists \(\Theta_{\Xi, \thyp{x}{A}{q}}\),
    \(\csplits{\Theta}{\Theta_\Delta}{\Theta_{\Xi, \thyp{x}{A}{q}}}\),
    \(\csplits{\Theta_{\Xi, \thyp{x}{A}{q}}}{\Theta_\Xi}{\Theta_x}\). Using the
    fact that \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\), we can conclude by
    \rle{subst-cons} that \(\issubst{\gamma}{\Theta_{\Xi, \thyp{x}{A}{q}}}{\Xi,
    \thyp{x}{A}{q}}\); the desired result hence follows.
    \item \rle{split-dup}: since \(\tlin{A}{\trel}\), it follows by
    \rle{subst-cons} that \(\tlin{\Theta_x}{\trel}\), and hence that
    \(\csplits{\Theta_x}{\Theta_x}{\Theta_x}\). Therefore, by split association,
    we have that there exist \(\Theta_{\Delta, \thyp{x}{A}{q}}\), \(\Theta_{\Xi,
    \thyp{x}{A}{q}}\) such that \(\csplits{\Theta}{\Theta_{\Delta,
    \thyp{x}{A}{q}}}{\Theta_{\Xi, \thyp{x}{A}{q}}}\). Using the facts that
    \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\),
    \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\), we can conclude by \rle{subst-cons}
    that \(\issubst{\gamma}{\Theta_{\Delta, \thyp{x}{A}{q}}}{\Delta,
    \thyp{x}{A}{q}}\), \(\issubst{\gamma}{\Theta_{\Xi, \thyp{x}{A}{q}}}{\Xi,
    \thyp{x}{A}{q}}\); the desired result hence follows.
    \item \rle{split-drop}: since \(\tlin{A}{\taff}\), it follows by
    \rle{subst-cons} that \(\tlin{\Theta_x}{\taff}\), and hence that
    \(\cwk{\Theta_x}{\cdot}\). Therefore, by split association,
    \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\), giving the desired result.
  \end{itemize}
\end{proof}

\begin{lemma}[Substitution] 
  Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
  \begin{itemize}
    \item If \(\hasty{\Gamma}{p}{a}{A}{q}\), then \(\hasty{\Theta}{p}{[\gamma]a}{A}{q}\)
    \item If \(\haslb{\Gamma}{t}{\ms{L}}\), then there exists \(\ms{K}\) such that \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) and \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\)
    \item If \(\lhaslb{\ms{W}}{L}{\ms{L}}\), then there exists \(\ms{K}, \ms{W}'\) such that 
    \begin{itemize}
      \item \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) 
      \item \(\lbsubst{\substlbs{\gamma}{\ms{W}'}}{\ms{W}}{\ms{W}'}\) 
      \item \(\lhaslb{\ms{W}'}{[\gamma]L}{\ms{K}}\)
    \end{itemize}
  \end{itemize}
\end{lemma}

\begin{proof} \
  \TODO{this}
\end{proof}

\begin{lemma}[Label Substitution] 
  Given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), 
  \begin{itemize}
    \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), \(\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}\)
    \item For all \(\lhaslb{\ms{W}}{L}{\ms{L}}\), \(\lhaslb{\ms{W}}{[\mc{L}]L}{\ms{K}}\)
  \end{itemize}
\end{lemma}

\begin{proof} \
  \TODO{this}
\end{proof}

\begin{lemma}[Semantic Upgrade]
  For all \(\Gamma, a, A\), if \(\hasty{\Gamma}{1}{a}{A}{q}\), then
  \[\dnt{\hasty{\Gamma}{0}{a}{A}{q}} = \upg{\dnt{\hasty{\Gamma}{1}{a}{A}{q}}}{}\]
\end{lemma}

\begin{proof} \
  \TODO{this}
\end{proof}

\begin{lemma}[Semantic Weakening] 
  If \(\cwk{\Gamma}{\Delta}\), then 
  \begin{itemize}
    \item For all \(\hasty{\Delta}{p}{a}{A}{q}\),
    \[
      \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
      ; \dnt{\hasty{\Delta}{p}{a}{A}{q}}
      = \dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
    \]
    \item For all \(\haslb{\Delta}{t}{\ms{L}}\),
    \[
      \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
      ; \dnt{\haslb{\Delta}{t}{\ms{L}}}
      = \dnt{\haslb{\Gamma}{t}{\ms{L}}}  
    \]
  \end{itemize}

  Similarly, if \(\lwk{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma}{t}{\ms{L}}\), then
  \[
    \dnt{\haslb{\Gamma}{t}{\ms{L}}}
    ; \dnt{\lwk{\ms{L}}{\ms{K}}}
    = \dnt{\haslb{\Gamma}{t}{\ms{K}}}
  \]
\end{lemma}

\begin{proof} \
  \TODO{this}
\end{proof}

\begin{lemma}[Substitution Coherence] \
  \begin{itemize}
    \item Given any two derivations \(D_1: \issubst{\gamma}{\Theta}{\Gamma}\), \(D_2: \issubst{\gamma'}{\Theta}{\Gamma}\) such that \(\forall x \in \Theta, \gamma(x) = \gamma'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
    \item Given any two derivations \(D_1: \lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), \(D_2: \lbsubst{\mc{L}'}{\ms{L}}{\ms{K}}\) such that \(\forall \lbl{\ell} \in \ms{L}, \mc{L}(x) = \mc{L}'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
  \end{itemize}
\end{lemma}

\begin{proof} \
  \TODO{this}
\end{proof}

\begin{lemma}[Semantic Substitution Splitting] 
  Given \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\issubst{\gamma}{\Delta}{\Xi}\),
  if \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\) and \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\), \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\), then
  \[
    \dnt{\issubst{\gamma}{\Theta}{\Gamma}};\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
    = \dnt{\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}};
      \dnt{\issubst{\gamma}{\Theta_\Delta}{\Delta}}
      \otimes \dnt{\issubst{\gamma}{\Theta_\Xi}{\Xi}}
  \]
\end{lemma}

\begin{proof} \
  \TODO{this}
\end{proof}

>>>>

\begin{figure}
  \begin{center}        
    \begingroup
    \renewcommand{\arraystretch}{1.5}
    \setlength{\tabcolsep}{2em}
    \begin{tabular}{rl}
        \multicolumn{1}{c}{Judgment} & \multicolumn{1}{c}{Meaning} \\ \hline
        \(\issubst{\gamma}{\Theta}{\Gamma}\) &
        \(\gamma\) is a substitution taking \(\Theta\) to \(\Gamma\) \\
        \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\) &
        \(\mc{L}\) is a label substitution taking \(\ms{L}\) to \(\ms{K}\)
    \end{tabular}
    \endgroup
  \end{center}
  \caption{Metatheoretic typing judgements for \isotopessa}
  \Description{Metatheoretic typing judgements for isotope-SSA}
  \label{fig:ssa-meta-judgements}
\end{figure}

@@

We begin with some basic definitions:
\begin{definition}[Binoidal Category] 
  A \textbf{binoidal category} \(\mc{C}\) is a category equipped with a binary
  operation on objects \(\otimes: \mc{C} \times \mc{C} \to \mc{C}\) which is
  independently functorial in each argument (i.e., for each object \(A\), \(A
  \otimes -\) and \(- \otimes A\) are functors).
\end{definition}
Given \(f: \mc{C}(A, B)\), \(g: \mc{C}(A', B')\) in a binoidal category, we
write \(f \ltimes g = f \otimes A';B \otimes g\) and \(f \rtimes g = A \otimes
g;f \otimes B'\); we say \(f\) is \textbf{central} if, for all \(g\), \(f
\ltimes g = f \rtimes g\) and \(g \ltimes f = g \rtimes f\), in which case we
simply write \(f \otimes g\), \(g \otimes f\) respectively.
\begin{definition}[Premonoidal Category] 
  A \textbf{premonoidal category} is a binoidal category with an identity object
  \(I: \mc{C}\) such that
  \begin{itemize}
    \item There exist central natural isomorphisms \(\lambda_A: I \otimes A
    \simeq A\) (the \textbf{left unitor}), \(\rho_A: A \otimes I \simeq A\) (the
    \textbf{right unitor})
    \item There exists a central isomorphism \(\alpha_{A, B, C}: (A \otimes B)
    \otimes C \simeq A \otimes (B \otimes C)\) (the \textbf{associator})
    \item Satisfying the \textbf{pentagon} and \textbf{triangle laws}; see e.g.
    \citet{maclane:71}.
  \end{itemize}
  We say a premonoidal category is \textbf{symmetric} if it is equipped with a
  central natural isomorphism \(\sigma_{A, B}: A \otimes B \simeq B \otimes A\)
  (the \textbf{symmetry}) satisfying the \textbf{hexagon law} (see
  \cite{maclane:71}). We say a premonoidal category is \textbf{monoidal} if
  every morphism is central.
\end{definition}
One issue with the notion of a premonoidal category is that it is unclear what
exactly a premonoidal functor should be: if we want them to compose, we would
like to preserve central morphisms, but trying to preserve \textit{all} central
morphisms is too strict a notion. We want, in essence, to be able to distinguish
morphisms that are truly "pure" from those which simply happen to be central. To
do that, we introduce the notion of an \textit{effectful category} as follows:
\begin{definition}[Effectful Category] 
  An \textbf{effectful category} \(\mc{V} \to \mc{C}\) is an identity on objects
  functor \(\upg{\cdot}{}: \mc{V} \to \mc{C}\) from a monoidal category
  \(\mc{V}\) to a premonoidal category \(\mc{C}\) which preserves all
  premonoidal structure (i.e. sends associators to associators, unitors to
  unitors, and tensor products to tensor products). It is \textbf{symmetric} if
  both \(\mc{V}\) and \(\mc{C}\) are, and \(\upg{\cdot}{}\) preserves the
  symmetry. A \textbf{Freyd category} is an effectful category where \(\mc{V}\)
  is Cartesian.
\end{definition}
Unfortunately, the notion of "purity" given by an effectful category only allows
us to perform relatively weak optimistions in the presence of effects. To
remedy this, we want to be able to make a finer distinction between morphisms
which are merely \textit{central} and those which are also \textit{relevant}
(i.e., copying the input and applying them twice is the same as applying them
once and copying the output) and/or \textit{affine} (i.e., applying them and
discarding the result is the same as never doing them at all). Doing so also
allows us to add the capability of \textit{distinguishing} between objects which
are affine, relevant, neither or both, enabling us to reason about substructural type systems categorically.

@@

Substitution only applies to central terms, because those are the only terms for
which substitution is sound. Furthermore, there are many other optimisations
that follow from the beta-eta theory of the language, as well as specific
properties of model-dependent operations (such as memory operations). In the
appendix, we state a general rewriting theorem which lifts equations between
terms into a full congruence. This lets us justify optimisations such as scalar
replacement of aggregates, eliminating redundant writes, reordering independent
memory operations, and memory-to-register transformations.

@@

% \section{Proofs}

% \subsection{Weakening and Substitution}

% \begin{theorem}[Weakening] \
%   \begin{itemize}
%     \item If \(\cwk{\Gamma}{\Delta}\) and \(q \subseteq r\),
%     \(\hasty{\Delta}{p}{a}{A}{r}\), then \(\hasty{\Gamma}{p}{a}{A}{q}\)
%     \item If \(\cwk{\Gamma}{\Delta}\), \(\haslb{\Delta}{t}{\ms{L}}\), then
%     \(\haslb{\Gamma}{t}{\ms{L}}\)
%     \item If \(\lwk{\ms{L}}{\ms{K}}\), \(\haslb{\Gamma}{t}{\ms{L}}\) then
%     \(\haslb{\Gamma}{t}{\ms{K}}\)
%   \end{itemize}
% \end{theorem}

% \begin{proof} \
%   \TODO{this}
% \end{proof}

% \begin{lemma}[Substitution Splitting] 
%   If \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\issubst{\gamma}{\Theta}{\Gamma}\),
%   then there exist \(\Theta_\Delta\), \(\Theta_\Xi\) such that:
%   \begin{itemize}
%     \item \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\)
%     \item \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\)
%     \item \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\)
%   \end{itemize}
% \end{lemma}

% \begin{proof} 
%   We proceed by induction on derivations \(\csplits{\Gamma}{\Delta}{\Xi}\). In
%   the \rle{split-nil} case, we have that \(\issubst{\gamma}{\cdot}{\cdot}\) by
%   \rle{subst-nil}. For all other cases, since \(\issubst{\gamma}{\Theta}{\Gamma,
%   \thyp{x}{A}{q}}\), we have by \rle{subst-cons} that
%   \(\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}\),
%   \(\issubst{\gamma}{\Theta_\Gamma}{\Gamma}\),
%   \(\hasty{\Theta_\Gamma}{1}{\gamma(x)}{A}{q}\). Since
%   \(\csplits{\Gamma}{\Delta}{\Xi}\), by induction, we have that
%   \(\csplits{\Theta_\Gamma}{\Theta_\Delta}{\Theta_\Xi}\) where \(\Theta_\Delta\)
%   and \(\Theta_\Xi\) have the desired properties above.
%   \begin{itemize}
%     \item \rle{split-left}: by left split association, we have that there exists
%     \(\Theta_{\Delta, \thyp{x}{A}{q}}\), \(\csplits{\Theta}{\Theta_{\Delta,
%     \thyp{x}{A}{q}}}{\Theta_\Xi}\), \(\csplits{\Theta_{\Delta,
%     \thyp{x}{A}{q}}}{\Theta_\Delta}{\Theta_x}\). Using the fact that
%     \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\), we can conclude by
%     \rle{subst-cons} that \(\issubst{\gamma}{\Theta_{\Delta,
%     \thyp{x}{A}{q}}}{\Delta, \thyp{x}{A}{q}}\); the desired result hence
%     follows.
%     \item \rle{split-right}: by right split association, we have that there
%     exists \(\Theta_{\Xi, \thyp{x}{A}{q}}\),
%     \(\csplits{\Theta}{\Theta_\Delta}{\Theta_{\Xi, \thyp{x}{A}{q}}}\),
%     \(\csplits{\Theta_{\Xi, \thyp{x}{A}{q}}}{\Theta_\Xi}{\Theta_x}\). Using the
%     fact that \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\), we can conclude by
%     \rle{subst-cons} that \(\issubst{\gamma}{\Theta_{\Xi, \thyp{x}{A}{q}}}{\Xi,
%     \thyp{x}{A}{q}}\); the desired result hence follows.
%     \item \rle{split-dup}: since \(\tlin{A}{\trel}\), it follows by
%     \rle{subst-cons} that \(\tlin{\Theta_x}{\trel}\), and hence that
%     \(\csplits{\Theta_x}{\Theta_x}{\Theta_x}\). Therefore, by split association,
%     we have that there exist \(\Theta_{\Delta, \thyp{x}{A}{q}}\), \(\Theta_{\Xi,
%     \thyp{x}{A}{q}}\) such that \(\csplits{\Theta}{\Theta_{\Delta,
%     \thyp{x}{A}{q}}}{\Theta_{\Xi, \thyp{x}{A}{q}}}\). Using the facts that
%     \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\),
%     \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\), we can conclude by \rle{subst-cons}
%     that \(\issubst{\gamma}{\Theta_{\Delta, \thyp{x}{A}{q}}}{\Delta,
%     \thyp{x}{A}{q}}\), \(\issubst{\gamma}{\Theta_{\Xi, \thyp{x}{A}{q}}}{\Xi,
%     \thyp{x}{A}{q}}\); the desired result hence follows.
%     \item \rle{split-drop}: since \(\tlin{A}{\taff}\), it follows by
%     \rle{subst-cons} that \(\tlin{\Theta_x}{\taff}\), and hence that
%     \(\cwk{\Theta_x}{\cdot}\). Therefore, by split association,
%     \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\), giving the desired result.
%   \end{itemize}
% \end{proof}

% \begin{lemma}[Substitution] 
%   Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
%   \begin{itemize}
%     \item If \(\hasty{\Gamma}{p}{a}{A}{q}\), then \(\hasty{\Theta}{p}{[\gamma]a}{A}{q}\)
%     \item If \(\haslb{\Gamma}{t}{\ms{L}}\), then there exists \(\ms{K}\) such that \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) and \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\)
%     \item If \(\lhaslb{\ms{W}}{L}{\ms{L}}\), then there exists \(\ms{K}, \ms{W}'\) such that 
%     \begin{itemize}
%       \item \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) 
%       \item \(\lbsubst{\substlbs{\gamma}{\ms{W}'}}{\ms{W}}{\ms{W}'}\) 
%       \item \(\lhaslb{\ms{W}'}{[\gamma]L}{\ms{K}}\)
%     \end{itemize}
%   \end{itemize}
% \end{lemma}

% \begin{proof} \
%   \TODO{this}
% \end{proof}

% \begin{lemma}[Label Substitution] 
%   Given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), 
%   \begin{itemize}
%     \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), \(\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}\)
%     \item For all \(\lhaslb{\ms{W}}{L}{\ms{L}}\), \(\lhaslb{\ms{W}}{[\mc{L}]L}{\ms{K}}\)
%   \end{itemize}
% \end{lemma}

% \begin{proof} \
%   \TODO{this}
% \end{proof}

% \subsection{Semantic Metatheory}

% \begin{lemma}[Semantic Upgrade]
%   For all \(\Gamma, a, A\), if \(\hasty{\Gamma}{1}{a}{A}{q}\), then
%   \[\dnt{\hasty{\Gamma}{0}{a}{A}{q}} = \upg{\dnt{\hasty{\Gamma}{1}{a}{A}{q}}}{}\]
% \end{lemma}

% \begin{proof} \
%   \TODO{this}
% \end{proof}

% \begin{lemma}[Semantic Weakening] 
%   If \(\cwk{\Gamma}{\Delta}\), then 
%   \begin{itemize}
%     \item For all \(\hasty{\Delta}{p}{a}{A}{q}\),
%     \[
%       \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
%       ; \dnt{\hasty{\Delta}{p}{a}{A}{q}}
%       = \dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
%     \]
%     \item For all \(\haslb{\Delta}{t}{\ms{L}}\),
%     \[
%       \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
%       ; \dnt{\haslb{\Delta}{t}{\ms{L}}}
%       = \dnt{\haslb{\Gamma}{t}{\ms{L}}}  
%     \]
%   \end{itemize}

%   Similarly, if \(\lwk{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma}{t}{\ms{L}}\), then
%   \[
%     \dnt{\haslb{\Gamma}{t}{\ms{L}}}
%     ; \dnt{\lwk{\ms{L}}{\ms{K}}}
%     = \dnt{\haslb{\Gamma}{t}{\ms{K}}}
%   \]
% \end{lemma}

% \begin{proof} \
%   \TODO{this}
% \end{proof}

% \begin{lemma}[Substitution Coherence] \
%   \begin{itemize}
%     \item Given any two derivations \(D_1: \issubst{\gamma}{\Theta}{\Gamma}\), \(D_2: \issubst{\gamma'}{\Theta}{\Gamma}\) such that \(\forall x \in \Theta, \gamma(x) = \gamma'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
%     \item Given any two derivations \(D_1: \lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), \(D_2: \lbsubst{\mc{L}'}{\ms{L}}{\ms{K}}\) such that \(\forall \lbl{\ell} \in \ms{L}, \mc{L}(x) = \mc{L}'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
%   \end{itemize}
% \end{lemma}

% \begin{proof} \
%   \TODO{this}
% \end{proof}

% \begin{lemma}[Semantic Substitution Splitting] 
%   Given \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\issubst{\gamma}{\Delta}{\Xi}\),
%   if \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\) and \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\), \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\), then
%   \[
%     \dnt{\issubst{\gamma}{\Theta}{\Gamma}};\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
%     = \dnt{\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}};
%       \dnt{\issubst{\gamma}{\Theta_\Delta}{\Delta}}
%       \otimes \dnt{\issubst{\gamma}{\Theta_\Xi}{\Xi}}
%   \]
% \end{lemma}

% \begin{proof} \
%   \TODO{this}
% \end{proof}

% \begin{theorem}[Semantic Substitution] 
%   Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
%   \begin{itemize}
%     \item For all \(\hasty{\Gamma}{p}{a}{A}{q}\), 
%     \[
%       \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{p}
%       ;\dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
%       = \dnt{\hasty{\Theta}{p}{[\gamma]a}{A}{q}}
%     \]
%     \item For all 
%       \(\haslb{\Gamma}{t}{\ms{L}}\), 
%       \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\), 
%       \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\),
%     \[
%       \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{}
%       ; \dnt{\haslb{\Gamma}{t}{\ms{L}}}
%       = \dnt{\haslb{\Theta}{[\gamma]t}{\ms{K}}} 
%       ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}}
%     \]
%     \item For all 
%       \(\lhaslb{\ms{W}}{L}{\ms{L}}\), 
%       \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\),
%       \(\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}\),
%       \[
%         \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
%         ; \dnt{\lhaslb{\ms{W'}}{[\gamma]L}{\ms{K}}}
%         = \dnt{\lhaslb{\ms{W}}{L}{\ms{L}}}
%         ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}}
%         + \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
%       \]
%   \end{itemize}
%   Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\),
%   \begin{itemize} 
%     \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), we have
%     \[
%       \dnt{\haslb{\Gamma}{t}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
%       = \dnt{\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}}
%     \]
%     \item For all \(\lhaslb{\ms{W}_{\ms{L}}}{L}{\ms{L}}\), we have
%     \[
%       \dnt{\lhaslb{\ms{W}_{\ms{L}}}{L}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
%       + \dnt{\lbsubst{\mc{L}}{\ms{W}_{\ms{L}}}{\ms{W}_{\ms{K}}}}
%       = \dnt{\lbsubst{\mc{L}}{\ms{W}_{\ms{L}}}{\ms{W}_{\ms{K}}}};\dnt{\lhaslb{\ms{W}_{\ms{K}}}{[\mc{L}]L}{\ms{K}}}
%     \]
%     where all variables in \(\ms{W}\) not in \(\ms{L}\) are fresh w.r.t \(\mc{L}\).
%   \end{itemize}
% \end{theorem}

% \begin{proof} \
%   \TODO{this}
% \end{proof}
